Emergent dual scaling of riverine biodiversity

### 2021-06-02 manuscript_pnas.Rmd
edits: manuscript Rmd
modified to work in word format

---
title: "Emergent dual scaling of riverine biodiversity"
output: word_document
csl: pnas.csl
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
bibliography: references.bib
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
---

@@

```{=tex}
\begin{equation}
  \mu_{z,k} = \rho \mu_{z,k+1} + \epsilon_{k}
\end{equation}
```
where $\mu_{z,k}$ is the mean environmental value at longitudinal position $k$ ($k$ is the network distance from the outlet patch; $k = 1$ at the outlet), $\rho$ is the strength of spatial autocorrelation and $\epsilon_k$ is the local environmental noise that follows a normal distribution with a mean of zero and a standard deviation of $\sigma_l$. The parameter $\rho$ can take values of 0-1 with larger values indicating greater spatial autocorrelation. In this study, we set $\rho = 1$ to mimic strong spatial autocorrelation in riverine environments. At confluences, we took a weighted mean of environmental values given the relative size of upstream contributing area of joining tributaries:

```{=tex}
\begin{equation}
\begin{aligned}
  \mu_{z,k} &= \omega(\rho \mu_{z,k+1}^R + \epsilon_{k}^R) + (1-\omega)(\rho \mu_{z,k+1}^L + \epsilon_{k}^L) \\
  \omega &= \frac{A_{k+1}^R}{A_{k+1}^R + A_{k+1}^L}
\end{aligned}
\end{equation}
```
where $R$ and $L$ denote "right" and "left" branches, respectively, and $A_k$ is the number of upstream habitat patches at longitudinal position $k$ (akin to the upstream watershed area; $A_k = 1$ at the most upstream patch). The parameter $\omega$ takes values of 0-1 and represents the relative influence of the right tributary given the relative size of joining tributaries. With this expression, larger tributaries have a greater influence on the downstream environment, as observed in natural river networks [@bendaNetworkDynamicsHypothesis2004].

@@

```{=tex}
\begin{equation}
  N_{ix}(t + 1) \sim Poisson(n_{ix}(t) + I_{ix}(t) - E_{ix}(t))
\end{equation}
```
where $n_{ix}(t)$ is the expected number of individuals given the local community dynamics at time $t$, $I_{ix}(t)$ the expected number of immigrants to patch $x$, and $E_{ix}(t)$ the expected number of emigrants from patch $x$. The realized discrete number of individuals is drawn from a Poisson distribution to account for stochasticity in demographic and dispersal processes [@fronhoferClassicalMetapopulationDynamics2017; @thompsonProcessbasedMetacommunityFramework2020]. Local community dynamics are simulated based on the Beverton-Holt equation [@bevertonDynamicsExploitedFish1957]:

```{=tex}
\begin{equation}
  n_{ix}(t) = \frac{N_{ix}(t)r_{ix}(t)}{1 + \frac{r_{0,i}-1}{K_{x}}\sum_{j=1}^S{\alpha_{ij}N_{jx}(t)}}
\end{equation}
```
where $r_{ix}(t)$ is the reproductive rate of species $i$ given the environmental condition at patch $x$ and time $t$, $r_{0,i}$ the maximum reproductive rate of species $i$, $K_{x}$ the carrying capacity at patch $x$, $\alpha_{ij}$ the competition coefficient between species $i$ and $j$, and $S$ the number of species in a metacommunity. $K_x$ was expressed as a function of the number of upstream habitat patches $A_x$ ($K_x = 100A_x$) to represent increasing habitat size with increasing upstream watershed area [@teruiThreeEcologicalFactors2016; @rahelFishAssemblagesHabitat1991]. The parameter $\alpha_{ij}$ represents the strength of interspecific competition relative to that of intraspecific competition: interspecific competition is greater than intraspecific competition if $\alpha_{ij}$ \> 1 (intraspecific competition coefficient was set constant as $\alpha_{ii} = 1$). $\alpha_{ij}$ was drawn randomly from a uniform distribution as $\alpha_{ij} \sim Unif(0, \alpha_{max})$. The density-independent reproductive rate $r_{ix}(t)$ is affected by abiotic environments (non-consumable) and determined by the following Gaussian function:

```{=tex}
\begin{equation}
  r_{ix}(t) = \delta~r_{0,i}~exp[{-\frac{\{\mu_i-z_x(t)\}^2}{2\sigma^2_{niche,i}}}]
\end{equation}
```
where $\mu_i$ is the optimal environmental value for species $i$, $z_x(t)$ the environmental value at patch $x$ and time $t$, and $\sigma_{niche,i}$ the niche width of species $i$. We introduced the cost of a wide niche by multiplying $\delta$ [@chaianunpornEvolutionaryResponsesClimate2015]:

```{=tex}
\begin{equation}
  \delta = exp(-\frac{\sigma_{niche,i}^2}{2\nu^2})
\end{equation}
```
Smaller values of $\nu$ imply greater costs of having a wider niche (i.e., decreased maximum reproductive rate) and reduce the realized reproductive rate $r_{ix}(t)$ near the niche optimum.

@@

```{=tex}
\begin{equation}
  \Omega_{xy} = \sigma_z^2~exp(-\phi d_{xy})
\end{equation}
```
$\Omega_{xy}$ denotes the temporal covariance of environmental conditions between patch $x$ and $y$, which is assumed to decay exponentially with increasing distance between the patches $d_{xy}$ (the shortest network path measured as the number of habitat patches). The parameter $\phi$ determines the degree of the distance decay of environmental correlates.

@@

```{=tex}
\begin{equation}
  \xi_{ix}(t) = \frac{\sum_{y,y \neq x}E_{iy}(t)~exp(-\theta d_{xy})}{\sum_{x} \sum_{y,y \neq x}E_{iy}(t)~exp(-\theta d_{xy})}
\end{equation}
```
The parameter $\theta$ regulates the dispersal distance of a species that follows an exponential distribution ($\theta^{-1}$ corresponds to the expected dispersal distance). The expected number of immigrants is calculated as $I_{ix}(t) = \xi_{ix}(t)\sum_x^{N_p} E_{ix}$.

@@

```{=tex}
\begin{equation}
  \mu_w = \xi_0 + \xi_1~log_{10}~A_w + \xi_2~log_{10}~P_{b,w} + \xi_3~region_w + \sum_k \xi_k x_{k,w}
\end{equation}
```
where $\xi_k$ are the intercept and regression coefficients for linear predictors. In an ordinary scale, the global model represents a dual scaling law as $E(y) = cA^{\xi_1}P_b^{\xi_2}$, where $c = 10^{\xi_0 + \xi_3region + \sum_k \xi_k x_{k}}$. Meanwhile, the region-specific model assumes the effects of watershed area and branching probability differ between regions (i.e., interactions between region and watershed area/branching probability).

```{=tex}
\begin{equation}
\begin{aligned}
  \mu_w =& \xi_0 + \xi_1~log_{10}~A_w + \xi_2~log_{10}~P_{b,w} + \xi_3~region_w +\\
  &\xi_4~log_{10}~A_w \cdot region_w + \xi_5~log_{10}~P_{b,w} \cdot region_w + \sum_k \xi_k x_{k,w}
\end{aligned}
\end{equation}
```
This is equivalent to:

```{=tex}
\begin{equation}
\begin{aligned}
  \mu_w =& \xi_0 + (\xi_1 + \xi_4~region_w)~log_{10}~A_w + (\xi_2 + \xi_5~region_w)~log_{10}~P_{b,w} + \xi_3~region_w + \sum_k \xi_k x_{k,w}
\end{aligned}
\end{equation}
```
Therefore, in an ordinary scale, the model assumes different scaling exponents between regions as $E(y) = cA^{\xi_1 + \xi_4region}P_b^{\xi_2 + \xi_5region}$. All explanatory variables except watershed area, branching probability, and region were standardized to a mean of zero and a standard deviation of one before the analysis.

@@

```{=tex}
\begin{equation}
  Bayes~factor = exp(\Delta BIC_{10}/2)
\end{equation}
```
where $\Delta BIC_{10} = BIC(M1) - BIC(M0)$.

>>>>

---
title: "Emergent dual scaling of riverine biodiversity"
output: word_document
csl: pnas.csl
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
bibliography: references.bib
---

@@

$$
\begin{aligned}
\mu_{z,k} = \rho \mu_{z,k+1} + \epsilon_{k} &\text{(1)}
\end{aligned}
$$

where $\mu_{z,k}$ is the mean environmental value at longitudinal position $k$ ($k$ is the network distance from the outlet patch; $k = 1$ at the outlet), $\rho$ is the strength of spatial autocorrelation and $\epsilon_k$ is the local environmental noise that follows a normal distribution with a mean of zero and a standard deviation of $\sigma_l$. The parameter $\rho$ can take values of 0-1 with larger values indicating greater spatial autocorrelation. In this study, we set $\rho = 1$ to mimic strong spatial autocorrelation in riverine environments. At confluences, we took a weighted mean of environmental values given the relative size of upstream contributing area of joining tributaries:

$$
\begin{aligned}
\mu_{z,k} &= \omega(\rho \mu_{z,k+1}^R + \epsilon_{k}^R) + (1-\omega)(\rho \mu_{z,k+1}^L + \epsilon_{k}^L) &&\\
\omega &= \frac{A_{k+1}^R}{A_{k+1}^R + A_{k+1}^L} &&\text{(2)}
\end{aligned}
$$

where $R$ and $L$ denote "right" and "left" branches, respectively, and $A_k$ is the number of upstream habitat patches at longitudinal position $k$ (akin to the upstream watershed area; $A_k = 1$ at the most upstream patch). The parameter $\omega$ takes values of 0-1 and represents the relative influence of the right tributary given the relative size of joining tributaries. With this expression, larger tributaries have a greater influence on the downstream environment, as observed in natural river networks [@bendaNetworkDynamicsHypothesis2004].

@@

$$
\begin{aligned}
N_{ix}(t + 1) \sim Poisson(n_{ix}(t) + I_{ix}(t) - E_{ix}(t)) &\text{(3)}
\end{aligned}
$$

where $n_{ix}(t)$ is the expected number of individuals given the local community dynamics at time $t$, $I_{ix}(t)$ the expected number of immigrants to patch $x$, and $E_{ix}(t)$ the expected number of emigrants from patch $x$. The realized discrete number of individuals is drawn from a Poisson distribution to account for stochasticity in demographic and dispersal processes [@fronhoferClassicalMetapopulationDynamics2017; @thompsonProcessbasedMetacommunityFramework2020]. Local community dynamics are simulated based on the Beverton-Holt equation [@bevertonDynamicsExploitedFish1957]:

$$
\begin{aligned}
n_{ix}(t) = \frac{N_{ix}(t)r_{ix}(t)}{1 + \frac{r_{0,i}-1}{K_{x}}\sum_{j=1}^S{\alpha_{ij}N_{jx}(t)}} &\text{(4)}
\end{aligned}
$$

where $r_{ix}(t)$ is the reproductive rate of species $i$ given the environmental condition at patch $x$ and time $t$, $r_{0,i}$ the maximum reproductive rate of species $i$, $K_{x}$ the carrying capacity at patch $x$, $\alpha_{ij}$ the competition coefficient between species $i$ and $j$, and $S$ the number of species in a metacommunity. $K_x$ was expressed as a function of the number of upstream habitat patches $A_x$ ($K_x = 100A_x$) to represent increasing habitat size with increasing upstream watershed area [@teruiThreeEcologicalFactors2016; @rahelFishAssemblagesHabitat1991]. The parameter $\alpha_{ij}$ represents the strength of interspecific competition relative to that of intraspecific competition: interspecific competition is greater than intraspecific competition if $\alpha_{ij}$ \> 1 (intraspecific competition coefficient was set constant as $\alpha_{ii} = 1$). $\alpha_{ij}$ was drawn randomly from a uniform distribution as $\alpha_{ij} \sim Unif(0, \alpha_{max})$. The density-independent reproductive rate $r_{ix}(t)$ is affected by abiotic environments (non-consumable) and determined by the following Gaussian function:

$$
\begin{aligned}
r_{ix}(t) = \delta~r_{0,i}~exp[{-\frac{\{\mu_i-z_x(t)\}^2}{2\sigma^2_{niche,i}}}] &\text{(5)}
\end{aligned}
$$

where $\mu_i$ is the optimal environmental value for species $i$, $z_x(t)$ the environmental value at patch $x$ and time $t$, and $\sigma_{niche,i}$ the niche width of species $i$. We introduced the cost of a wide niche by multiplying $\delta$ [@chaianunpornEvolutionaryResponsesClimate2015]:

$$
\begin{aligned}
\delta = exp(-\frac{\sigma_{niche,i}^2}{2\nu^2}) &\text{(6)}
\end{aligned}
$$

Smaller values of $\nu$ imply greater costs of having a wider niche (i.e., decreased maximum reproductive rate) and reduce the realized reproductive rate $r_{ix}(t)$ near the niche optimum.

@@

$$
\begin{aligned}
\Omega_{xy} = \sigma_z^2~exp(-\phi d_{xy}) &\text{(7)}
\end{aligned}
$$

$\Omega_{xy}$ denotes the temporal covariance of environmental conditions between patch $x$ and $y$, which is assumed to decay exponentially with increasing distance between the patches $d_{xy}$ (the shortest network path measured as the number of habitat patches). The parameter $\phi$ determines the degree of the distance decay of environmental correlates.

@@

$$
\begin{aligned}
\xi_{ix}(t) = \frac{\sum_{y,y \neq x}E_{iy}(t)~exp(-\theta d_{xy})}{\sum_{x} \sum_{y,y \neq x}E_{iy}(t)~exp(-\theta d_{xy})} &\text{(8)}
\end{aligned}
$$

The parameter $\theta$ regulates the dispersal distance of a species that follows an exponential distribution ($\theta^{-1}$ corresponds to the expected dispersal distance). The expected number of immigrants is calculated as $I_{ix}(t) = \xi_{ix}(t)\sum_x^{N_p} E_{ix}$.

@@

$$
\begin{aligned}
\mu_w = \xi_0 + \xi_1~log_{10}~A_w + \xi_2~log_{10}~P_{b,w} + \xi_3~region_w + \sum_k \xi_k x_{k,w} &\text{(9)}
\end{aligned}
$$

where $\xi_k$ are the intercept and regression coefficients for linear predictors. In an ordinary scale, the global model represents a dual scaling law as $E(y) = cA^{\xi_1}P_b^{\xi_2}$, where $c = 10^{\xi_0 + \xi_3region + \sum_k \xi_k x_{k}}$. Meanwhile, the region-specific model assumes the effects of watershed area and branching probability differ between regions (i.e., interactions between region and watershed area/branching probability).

$$
\begin{aligned}
  \mu_w =& \xi_0 + \xi_1~log_{10}~A_w + \xi_2~log_{10}~P_{b,w} + \xi_3~region_w + &&\\
  &\xi_4~log_{10}~A_w \cdot region_w + \xi_5~log_{10}~P_{b,w} \cdot region_w + \sum_k \xi_k x_{k,w}
&&\text{(10)}
\end{aligned}
$$

This is equivalent to:

$$
\begin{aligned}
  \mu_w =& \xi_0 + (\xi_1 + \xi_4~region_w)~log_{10}~A_w + (\xi_2 + \xi_5~region_w)~log_{10}~P_{b,w} + &&\\
  &\xi_3~region_w + \sum_k \xi_k x_{k,w}
&&\text{(11)}
\end{aligned}
$$

Therefore, in an ordinary scale, the model assumes different scaling exponents between regions as $E(y) = cA^{\xi_1 + \xi_4region}P_b^{\xi_2 + \xi_5region}$. All explanatory variables except watershed area, branching probability, and region were standardized to a mean of zero and a standard deviation of one before the analysis.

@@

$$
\begin{aligned}
Bayes~factor = exp(\Delta BIC_{10}/2) &\text{(12)}
\end{aligned}
$$

where $\Delta BIC_{10} = BIC(M1) - BIC(M0)$.
